Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-10-21 Paper/ms.tex
Theta

\begin{abstract}
Inferring the number of planets $N$ in an exoplanetary system from radial velocity
(RV) data is a challenging task. Typically, the posterior distribution for the
planet parameters, such as orbital periods,
is multimodal and may have other awkward features. In
addition, when $N$ is unknown, the marginal likelihood (or evidence), as a
function of $N$, is required. In this paper we propose an alternative
approach using a trans-dimensional Markov Chain Monte Carlo method within
Nested Sampling. The posterior distribution for $N$ can be obtained with a
single run, which takes $\sim$ 15 minutes on a typical modern desktop computer.
We apply the method to $\nu$ Oph and Gliese 581, finding moderate evidence
(posterior probability of existence = 70\%) for a third planet around
$\nu$ Oph with an orbital period of 75.7 $\pm$ 1.8 days. The results also
suggest Gliese 581 hosts many (7-15) planets (or causes of other periodic
signals), but only four have well determined orbital periods. 
\end{abstract}

@@

Consider a test particle moving in the $x$-$y$ plane under the influence of a
point mass at the origin. The motion of the test particle represents the
reflex motion of the host star orbiting around the center of mass of the
system. The equations of motion for the particle are:
\begin{eqnarray}
\frac{d^2x}{dt^2} &=& -\frac{x}{r^3} \\
\frac{d^2y}{dt^2} &=& -\frac{y}{r^3} \\
\end{eqnarray}
where $r = \sqrt{x^2 + y^2}$. These ODEs can be solved for any given set of
initial conditions for $x$, $y$, $\frac{dx}{dt}$, and $\frac{dy}{dt}$ to yield
the shape of the orbit. To simulate orbits
of different ellipticities, we set the initial position to $(1, 0)$, and
the initial velocity to $(0, v)$ where $v \in [0.4, 1]$.
If $v=1$, the orbit is circular and as $v$ decreases the orbit becomes more
elliptical. For trial values of $v$ ranging from 0.4 to 1 in steps of 0.005,
we simulated an orbit using a leapfrog integrator, and saved the
velocities $\frac{dx}{dt}$ and $\frac{dy}{dt}$ as a function of time to disk.
These saved orbits were used as a lookup table for constructing the expected
signal $y(t)$ due to a single planet.
Because of the initial conditions, the simulated orbits were all horizontally
aligned. If the observer is located on the $x$-axis a large distance
from the origin, they will measure $m(t) = \dot{x}(t)$. However, if the
observer is located at

>>>>

\begin{abstract}
Inferring the number of planets $N$ in an exoplanetary system from radial velocity
(RV) data is a challenging task. Typically, the posterior distribution for the
planet parameters, such as orbital periods,
is multimodal and may have other awkward features. In
addition, when $N$ is unknown, the marginal likelihood (or evidence), as a
function of $N$, is required. In this paper we propose an alternative
approach using a trans-dimensional Markov Chain Monte Carlo method within
Nested Sampling. The posterior distribution for $N$ can be obtained with a
single run, which takes $\sim$ 15 minutes on a typical modern desktop computer.
We apply the method to $\nu$ Oph and Gliese 581, finding moderate evidence
(posterior probability of existence = 70\%) for a third planet around
$\nu$ Oph with an orbital period of 75.7 $\pm$ 1.8 days. The results also
suggest Gliese 581 hosts many (7-15) planets (or other causes of other periodic
signals), but only four have well determined orbital periods. 
\end{abstract}

@@

Consider a test particle moving in the $x$-$y$ plane under the influence of a
point mass at the origin. The motion of the test particle represents the
reflex motion of the host star orbiting around the center of mass of the
system. The equations of motion for the particle are:
\begin{eqnarray}
\frac{d^2x}{dt^2} &=& -\frac{x}{r^3} \\
\frac{d^2y}{dt^2} &=& -\frac{y}{r^3} \\
\end{eqnarray}
where $r = \sqrt{x^2 + y^2}$. These ODEs can be solved for any given set of
initial conditions for $x$, $y$, $\frac{dx}{dt}$, and $\frac{dy}{dt}$ to yield
the shape of the orbit. To simulate orbits
of different ellipticities, we set the initial position to $(1, 0)$, and
the initial velocity to $(0, v)$ where $v \in [0.4, 1]$.
If $v=1$, the orbit is circular and as $v$ decreases the orbit becomes more
elliptical. For trial values of $v$ ranging from 0.4 to 1 in steps of 0.005,
we simulated an orbit using a leapfrog integrator, and saved the
velocities $\frac{dx}{dt}$ and $\frac{dy}{dt}$ as a function of time to disk.
These saved orbits were used as a lookup table for constructing the expected
signal $y(t)$ due to a single planet.
Because of the initial conditions, the simulated orbits were all horizontally
aligned. If the observer is located on the $x$-axis a large distance
from the origin, they will measure $m(t) = \dot{x}(t)$. However, if the
observer is located at an angle $\theta$ with respect to the $x$-axis, then
the radial velocity measured will instead be
$m(t) = \cos(\theta)\dot{x}(t) + \sin(\theta)\dot{y}(t)$.
Since the our orientation with respect to the orbits is unknown, each planet
requires a ``viewing angle'' parameter $\theta$.
